Supporting information: Tail associations and their impact on extinction risk

### 2019-11-03 MainText.Rmd
working on sup mat

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between the values of an environmental variable in different
locations, $X_i(t)$ and $X_j(t)$, using standard methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used are recommended instead.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even through Pearson correlation
appears to inducate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends on both marginal information and association information.
It is possible that recently observed changes in environmental
[e.g., @Black2018] and biological [e.g., @Koenig2016] synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of  
variables, rather than reflecting true changes in association.
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, and how such 
associations may be changing.

>>>>

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between the values of an environmental variable in different
locations, $X_i(t)$ and $X_j(t)$, using standard methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used are recommended instead.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even through Pearson correlation
appears to inducate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends on both marginal information and association information.
It is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of  
variables, rather than reflecting true changes in association.
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, and how such 
associations may be changing.

### 2019-11-03 SuppMat.Rmd
edits to results
working on discussion
working on sup mat

dt4_2<-tibble("$r_{min}$" = c("0",
                        "1",
                        "1",
                        "$1+\\exp(-a)$",
                        "0",
                        "0"
))

@@

<!--needs editing, I just dumped old text here-->

Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. Copulas separate 
the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables.

<!--# Algorithm to generate extreme tail association  \label{extr_tail_dep}-->

For extreme tail association algorithm, a $N \times n$ matrix, M was generated of independent 
standard-normal random draws (in general). <!--where $N = numsteps(=25) \times numsims(=10000)$ : only for 2d case-->
Then, each row of M was identified for a first set of modifications
randomly and independently with a 50\% chance; for those rows identified, every
element of the row was replaced with the absolute
value of the first element of the row (including the first element itself).
For other rows, every element in the row was replaced by the negative of 
its own absolute value. This produces right-tail dependent
matrix, normally distributed for each column. If left-tail 
dependent noise is desired, take the negative of the entire matrix.
Our [code](https://github.com/sghosh89/ERC/blob/master/ExtremeTailDep.R) for
generating extreme tail association matrix is flexible enough, so that 
matrix M can be generated initially with independent random draws from a normal 
distribution with a specified mean and standard deviation.

<!--some stuff that used to be in the main text-->

The moderate tail association was modeled by using the 
concept of "copula", a statistical tool to model 
dependencies. Appendix \ref{SM-intro_copula} summarized
a brief introduction of "copulas" and their relation to tail associations, symmetries 
with examples of available copula families 
from `R` packages (`copula` and `VineCopula`). We used bivariate Clayton 
copula to model moderate left-tail and Survival Clayton copula to model 
moderate right-tail association in the vital rates. Both copulae had the same Spearman 
correlation ($\rho =$ `r srho_common`) but opposite association in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail associated bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iRho` function from `copula` package so that both copula had
same Spearman correlation ($\rho =$ `r srho_common`). Entries within each column of
matrix $M$, then became independent of each other but 
entries between columns had a specific tail association
with positive Spearman correlation.

<!--some more stuff from the main text-->

On a similar note, we generated extreme left- (or right-) tail dependent matrix (M)
from standard iid (Appendix \ref{SM-extr_tail_dep}) and made the marginals uniform so that
it could mimic an extreme tail dependent copula. We made the first column of matrix 
$M$ as gamma ($\Gamma(k,\theta)$) distributed 
with shape parameter $k = 2$ and scale parameter $\theta = 1$.
This distribution represented fecundity of the adult population, $f_a \geq 0$.
For a given $\theta$, increasing $k$ would shift the distribution towards
higher fecundity and reduce the extinction risk if the other parameters
remain unchanged. We took the second column of $M$ and applied a 
beta ($\mathcal{B}(\alpha,\beta)$) transformation. $\alpha =3$ and
$\beta = 2$ were the shape parameters that could control the distribution 
of the survival rates of eggs ($s_e$) ranging in between 0 to 1. 
For each $25$ timesteps from each of $10000$ simulations,
we incorporated pairwise entries from each row of matrix $M$ in the expression for 
vital rates (fecundity $f_a$ and survival rates $s_e$, Eq. \ref{eq_vitalrates}) of 
the stochastic projrction matrix in model (\ref{eq_age_str}).

<!--yet more stuff from the main text-->

For moderately left-tail associated 
noises we drew points from a $n$ dimensional 
Clayton copula with same Spearman correlation ($\rho =$ `r srho_common`) between
any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail associated noises we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail association. We generated
extreme left- or right-tail associated noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}.

@@

First, we will briefly discuss about the stability analysis for a difference 
equation which is frequently used in population ecology model. Let's say, $P(t+1) = F(P(t),r)$
represents the model dynamics of a single population $P$ which progress recursively via 
a nonlinear density-dependence function $F$ with discrete time variable $t$. $r$ is the model parameter 
which plays major role in assessing the model's dynamical stability. Fixed-point analysis can
be used to get information about this type of nonlinear difference equation. The idea is to
employ the concept that the system at fixed point (or equilibrium $P_e$) does not change with time 
(i.e. $P(t+1) = P(t)$). Given a tiny perturbation from its equilibrium $P_e$, if the system 
returns to its initial state, then $P_e$ is called to be a stable equilibrium, otherwise unstable.
Mathematically, from standard linear stability analysis we can say an equilibrium is locally 
asymptotically stable if $| {F}' | < 1$, where ${F}' = \left( \frac{\mathrm{d}F }{\mathrm{d} P} \right) _{P(t)=P_e}$.
If the approach towards its equilibrium is monotonic (when $0<{F}'<1$), the dynamics is said to be
in under-compensatory regime. An over-compensatory dynamics means the approach towards its stable equlibrium
is oscillatory ($-1<{F}'<0$). Following we will describe stability condition of the model dynamics considered for each 
patch of metapopulation models in the maintext (see Table \ref{MT-tab_model_summary}).

>>>>

dt4_2<-tibble("$r_{min}$" = c("0",
                        "1",
                        "1",
                        "1.90",
                        "0",
                        "0"
))

@@

<!--# Algorithm to generate extreme tail association  \label{extr_tail_dep}-->
Bivarite environmental noise with extreme left-tail association (as in Fig. 
\ref{fig_intro_cop}A) was generated using an algorithm originally presented 
by @Ghosh_copula (their Appendix S10), presented again here for convenience.
If $T$ time steps of noise were desired, step 1 was to generate a $T$ by $2$
matrix, $M$, of independent random numbers, each standard-normally distributed. 
Then, each row of $M$ was identified for modification randomly and
independently, with a 50\% chance. For those rows identified, every element
of the row was replaced by the negative of the absolute value of the first element
of the row, including the first element itself. For the other rows, every 
element in the row was replaced by its own absolute value. For noise
with extreme right-tail association (as in Fig. 
\ref{fig_intro_cop}D), take the negative of the entire matrix. For
both types of noise, every entry was transformed by the cummulative distribution
function of a standard normal distribution, to make each column uniformly
distributed.

<!--Spearman correlation of the extreme-tail-associated noise-->
The Spearman correlation of the extreme left- or right-tail associated 
noise was $0.875$, up to sampling variation.
<!--***DAN: fill this in-->

<!--moderate tail association noise-->
Bivarite environmental noise with moderate left-tail association (as in Fig. 
\ref{fig_intro_cop}B) was generated using a 2-dimensional Clayton copula. The statistical
theory of copulas has been introduced systematically in several texts 
(e.g., @nelsen2006_copula; @joe2014_dependence), and was briefly 
introduced by @Genest2007. Short, conceptual introductions aimed at ecological
audiences were provided by sections within the papers of @Anderson2018 and  
@Ghosh_copula. A 2-dimensional Clayton copula is equivalent to a particular bivariate 
probability density function with marginal distributions which are 
uniform on the interval $(0,1)$. The points on Fig. 
\ref{fig_intro_cop}B were independent draws from the associated random variable.
The functions `claytonCopula` and `rCopula` from the `copula` package for the 
R programming language were used. The parameter of the Clayton family was
chosen so that the Spearman correlation was $0.875$, the same as that of the
extreme left-tail associated noise. This was accomplished with the `iRho`
function of the `copula` package. Noise with moderate right-tail
association was generated in a similar way but using the 2-dimensional 
survival Clayton copula.

<!--Multivariate extreme tail association-->
Multivariate noise with extreme left-tail association was generated using a
very simlar algorithm to the bivariate case,
as follows. If $T$ time steps of noise of dimension $n$ were desired,
step 1 was to generate a $T$ by $n$
matrix, $M$, of independent random numbers, each standard-normally distributed.
Each row of $M$ was identified for modification randomly and
independently, with a 50\% chance. For those rows identified, every element
of the row was replaced by the negative of the absolute value of the first element
of the row, including the first element itself. For the other rows, every 
element in the row was replaced by its own absolute value. For noise
with extreme right-tail association, take the negative of the entire matrix. For
both types of noise, every entry was transformed by the cummulative distribution
function of a standard normal distribution, to make each column uniformly
distributed. 
Multivariate moderately tail associated noise of dimension $n$
was generated using $n$-dimensional Clayton or survival Clayton copulas.

@@

Let $P(t+1) = F(P(t),r)$ be a deterministic population-dynamic model, where 
$P(t)$ is the population at time $t$ and $r$ is a model parameter. Let $P_e$
be an equilibrium (which may depend on $r$), i.e., $F(P_e,r)=P_e$. $P_e$
is called a stable equilibrium if, when the population is perturbed slightly
from equilibrium, the system returns to the equilibrium value; $P_e$ is called
an unstable equilibrium otherwise.
From standard linear stability analysis, we say an equilibrium is locally 
asymptotically stable if $| {F}' | < 1$, where ${F}' = \left( \frac{\mathrm{d}F }{\mathrm{d} P} \right) _{P=P_e}$.
If $0<{F}'<1$, then the approach of the system to the equilibrium is monotonic,
and dynamics are called undercompensatory. If $-1<{F}'<0$, then the approach of the
system to equilibrium is oscillatory, and dynamics are called overcompensatory. 
We will now describe stability analyses of the deterministic 1-habitat-patch version
of each model of Table \ref{MT-tab_model_summary}.

<!--***DAN: got to here-->
